MR-PFU: A suite of R packages for energy conversion chain analysis

### 2023-07-28 paper.Rmd
Edits to paper.

Energy flowing through societal energy conversion chains (ECCs)
enables economic activity and facilitates human flourishing.
To understand economic growth and human well-being, 
the field of societal energy analysis (among other activities)
evaluates ECCs
from the primary stage 
(resources extracted from the environment, 
such as coal, oil, and natural gas) 
to the final stage
(energy purchased by consumers, 
such as refined petroleum and electricity),
to the useful stage
(energy desired by the end user, 
such as heat, motion, and light),
and sometimes to energy services
(such as thermal comfort, transport, and illumination).
Societal exergy analysis (SEA), 
an extension of energy analysis,
quantifies ECCs in exergy terms^[Exergy is 
the mechanical work potential of energy.].

We created a suite of `R` packages 
(some of which are generally useful and available through 
[CRAN][link-cran],
others of which are focused on SEA and available on GitHub)
to assist SEA practitioners as they analyze 
energy movement through society. 
The new packages enable analysis of any country in the world
across timespans of decades or longer.
In short, the packages enable, for the first time, scalable SEA.
We used the new packages to create the "PFU Database,"^[Technically speaking, 
we create data frames of matrices, 
not an SQL or similar database.] 
a new resource for the SEA community [@Marshall:2023aa].

@@

Historically, SEA practitioners have
analyzed the ECCs of individual countries
using linked spreadsheets,
often starting with primary- and final-stage data 
from the [IEA][link-ieaweeb]'s
world energy balances
[@Ayres:2003ec; @Serrenho:2014aa; @Brockway:2014aa; @Brockway:2015aa]. 
Data were stored in varying and inconsistent formats.
An early attempt to create a worldwide SEA database [@De-Stercke:2014]
estimated technical energy efficiencies 
of end-use machines by economic status in a country,
thereby precluding the use of the database for energy-economy studies.

The authors of the current paper and others in the field
wanted to expand SEA to cover all countries, but
the spreadsheet approach to SEA was
deemed not scalable.
A new approach to SEA was needed, 
one that scaled across all countries 
for many years without relying on economic data
to estimate technical efficiencies.

@@

The most important decision for the design of the suite of `R` packages
involves data format.
We authors are among those who
developed the Physical Supply-Use Table (PSUT) framework,
a matrix-based approach to describing energy flows 
("products" in PSUT terminology) 
from resource extraction to
processing stages
("industries" in PSUT terminology)
and, ultimately, 
to final demand [@Rocco:2016; @Guevara:2017; @Heun:2018; @Aramendia:2022tv].
We chose the PSUT framework as the data format for the `R` packages, 
because it succinctly describes an entire ECC
via a set of six matrices described in the following table.

@@

Further development followed from selection of the PSUT framework.
First, the matrices of the PSUT framework carry the challenge that 
different countries and years 
have varying energy carriers (products) and
varying energy conversion machines (industries),
meaning that PSUT matrices for different countries and years 
have differing sizes and differing row and column names.
To get around this challenge, we created the
[matsbyname][link-matsbyname] package [@Heun-matsbyname:2023]
which enables matrix mathematics 
that respects row and column names, 
inserting rows or columns of `0`s when needed.
Second,
we knew it would be convenient to perform _matrix_ mathematics 
as easily as _scalar_ mathematics
in `R` data frames
using the [tidyverse][link-tidyverse] syntax [@Wickham:2019]. 
We developed the [matsindf][link-matsindf] package [@Heun-matsindf:2023]
to enable this functionality.
Finally, manipulating row and column names proved to be a challenge, 
especially for matrices in PSUT data frames, so 
we developed the [RCLabels][link-rclabels] package [@Heun-RCLabels:2023]
to assist.
The table below summarizes these packages, 
all of which are generally useful and available on 
[CRAN][link-cran].

| Package | Purpose |
|:--------|:--------|
| [RCLabels][link-rclabels]     | Manipulates row and column names in [matsindf][link-matsindf] data frames |
| [matsbyname][link-matsbyname] | Performs matrix mathematics that respects row and column names |
| [matsindf][link-matsindf]     | Stores matrices in cells of a data frame, thereby enabling [tidyverse][link-tidyverse] syntax |

Four calculation steps are required to create a PFU database. 
Each step is assisted by a new `R` package.
First, the IEA's primary- and final-stage WEEB data must be converted
to the PSUT format, 
a task completed by the [IEATools][link-ieatools] package [@Heun-IEATools:2023]. 
Second, human and animal muscle work must be calculated from 
[International Labor Organization][link-ilo] (ILO) and
[Food and Agriculture Organization][link-fao] (FAO) data,
following the methodology of @Steenwyk:2022ww, 
using the [MWTools][link-mwtools] package [@Marshall:2023ab]. 
Third, the IEA's primary- and final-stage ECC data 
are extended to the useful stage by
(a) allocating final stage energy to end-use machines and
(b) multiplying allocated final energy by the 
final-to-useful efficiency of each machine.
This task is accomplished by the
[Recca][link-recca] [@Heun-Recca:2023]
and 
[PFUDatabase][link-pfudatabase] [@Heun-PFUDatabase:2023]
packages.
Fourth, ECCs must be converted from energy terms to exergy terms. 
This step is assisted by the [Recca][link-recca] package.

The steps to create the PSUT matrices for each country and each year
are accomplished by a
[targets][link-targets] [@Landau:2021aa]
computation pipeline
available in the [PFUDatabase][link-pfudatabase]
package.
A unique feature of the
[PFUDatabase][link-pfudatabase]
pipeline is an exemplar system that allows
analyses to proceed when allocation or efficiency data for a country are unavailable.
The calculation pipeline in the [PFUDatabase][link-pfudatabase] packge 
also allows allocation and efficiency data
for any country to be added at any time to improve the database.

@@

| Package | Purpose |
|:--------|:--------|
| [IEATools][link-ieatools]             | Converts IEA data to [matsindf][link-matsindf] format |
| [MWTools][link-mwtools]               | Converts [ILO][link-ilo] and [FAO][link-fao] data to human and animal muscle work in [matsindf][link-matsindf] format |
| [Recca][link-recca]                   | Performs `R` energy conversion chain analysis |
| [PFUSetup][link-pfusetup]             | Identifies input and output locations for the [PFUDatabase][link-pfudatabase] and [PFUAggDatabase][link-pfuaggdatabase] pipelines |
| [PFUDatabase][link-pfudatabase]       | A [targets][link-targets] pipeline to create a data frame of PSUT matrices |
| [PFUAggDatabase][link-pfuaggdatabase] | A [targets][link-targets] pipeline to aggregate PSUT matrices |

Input data for the PFU Database can be found in @Marshall:2023aa.
Access to the PFU Database can be obtained via correspondence 
with author [PEB](mailto:P.E.Brockway@leeds.ac.uk)^[Because the PFU Database contains
primary and final energy [IEA WEEB][link-ieaweeb] data, 
access to the PFU Database
is restricted to those who have access to [IEA WEEB][link-ieaweeb] data.].

>>>>

Energy flowing through societal energy conversion chains (ECCs)
enables economic activity and facilitates human flourishing.
To understand economic growth and human well-being, 
the field of societal energy analysis (among other activities)
evaluates ECCs
from the primary stage 
(resources extracted from the environment, 
such as coal, oil, natural gas, wind, and solar) 
to the final stage
(energy purchased by consumers, 
such as refined petroleum and electricity),
to the useful stage
(energy desired by the end user, 
such as heat, motion, and light),
and sometimes to energy services
(such as thermal comfort, transport, and illumination).
Societal _exergy_ analysis (SEA), 
an extension of societal _energy_ analysis,
quantifies ECCs in exergy terms^[Exergy is 
the mechanical work potential of energy.
See any number of web references for an exergy primer,
including @SciDirect:2023. 
**** Do we have a better primer? ****].

We created a suite of `R` packages 
to assist SEA practitioners to analyze 
energy movement through society. 
The new packages enable analysis of any country in the world
across timespans of decades or longer.
In short, the packages enable, for the first time, scalable SEA.
We used the new packages to create the MR-PFU database,^["MR-PFU" 
is an initialism for "Multi-Region
Primary, Final, Useful."
The MR-PFU database is multi-regional in the sense that
it contains many countries as well as 
continental and world aggregations.
Technically speaking, 
we create [matsindf][link-matsindf] data frames, 
not SQL or similar databases.] 
a new resource for the SEA community [@Marshall:2023aa].

@@

Historically, SEA practitioners have
analyzed the ECCs of individual countries
using linked spreadsheets,
often starting with primary- and final-stage data 
from the [IEA][link-ieaweeb]'s
world energy balances
[@Ayres:2003ec; @Serrenho:2014aa; @Brockway:2014aa; @Brockway:2015aa]. 
Data were stored in varying and inconsistent formats.
An early SEA database [@De-Stercke:2014]
estimated energy efficiencies 
of end-use machines by the economic status of countries,
thereby precluding its use for energy-economy studies.

The authors of the current paper and others in the field
wanted to expand SEA to cover all countries, but
the spreadsheet approach to SEA was
deemed not scalable.
A new approach to SEA was needed, 
one that scaled across all countries 
for many years without relying on economic data
to estimate machine efficiencies.

@@

The most important design decision for the suite of `R` packages
involved data format.
We authors are among those who
developed the Physical Supply-Use Table (PSUT) framework,
a matrix-based approach to describing energy flows 
("products" in PSUT terminology) 
from resource extraction to
processing stages
("industries" in PSUT terminology)
and, ultimately, 
to final demand [@Rocco:2016; @Guevara:2017; @Heun:2018; @Aramendia:2022tv].
We chose the PSUT framework as the data format for the `R` packages, 
because it succinctly describes an entire ECC
for one country and one year
via a set of six matrices
(the "RUVY" matrices).
See the following table.

@@

Further development followed from selecting the PSUT framework.
First, the RUVY matrices carry the challenge that 
different countries and years 
have varying energy carriers (products) and
varying energy conversion machines (industries),
meaning that RUVY matrices 
have differing row names, differing column names, and differing sizes.
To get around this challenge, we created the
[matsbyname][link-matsbyname] package [@Heun-matsbyname:2023]
which enables matrix mathematics 
that respects row and column names, 
inserting **0** vectors when needed.
Second,
wanted to be able to perform _matrix_ mathematics 
as easily as _scalar_ mathematics
in `R` data frames
using the [tidyverse][link-tidyverse] syntax [@Wickham:2019]. 
We developed the [matsindf][link-matsindf] package [@Heun-matsindf:2023]
to enable this functionality.
Finally, manipulating row and column names proved to be a challenge, 
especially for RUVY matrices in [matsindf][link-matsindf] data frames, so 
we developed the [RCLabels][link-rclabels] package [@Heun-RCLabels:2023]
to assist.
The table below summarizes these packages, 
all of which are generally useful and available on 
[CRAN][link-cran].

| Package | Purpose |
|:--------|:--------|
| [RCLabels][link-rclabels]     | Manipulates matrix row and column names in [matsindf][link-matsindf] data frames |
| [matsbyname][link-matsbyname] | Performs matrix mathematics that respects row and column names |
| [matsindf][link-matsindf]     | Stores matrices in cells of data frames, thereby enabling analysis with [tidyverse][link-tidyverse] syntax |

Broadly speaking, 
four calculation steps are required to create a PFU database. 
Each step is completed by one or more new `R` packages.
First, the IEA's primary- and final-stage WEEB data must be converted
to RUVY matrices for each country and year, 
a task accomplished by functions in the [IEATools][link-ieatools] package [@Heun-IEATools:2023]. 
Second, human and animal muscle work must be calculated from 
[International Labor Organization][link-ilo] (ILO) and
[Food and Agriculture Organization][link-fao] (FAO) data,
following the methodology of @Steenwyk:2022ww, 
using functions in the [MWTools][link-mwtools] package [@Marshall:2023ab]. 
Third, the IEA's primary- and final-stage ECC data 
must be extended to the useful stage by
(a) allocating final stage energy to end-use machines and
(b) multiplying allocated final energy by the 
final-to-useful efficiency of each machine.
This task is accomplished by functions in the
[Recca][link-recca] [@Heun-Recca:2023]
and 
[PFUDatabase][link-pfudatabase] [@Heun-PFUDatabase:2023]
packages.
Fourth, ECCs must be converted from energy terms to exergy terms. 
This step is performed by functions in the [Recca][link-recca] package.

The steps to create the PSUT matrices for each country and each year
are accomplished by a
[targets][link-targets] [@Landau:2021aa]
computation pipeline
available in the [PFUDatabase][link-pfudatabase]
package.
A unique feature of the
[PFUDatabase][link-pfudatabase]
pipeline is an exemplar system that allows
analyses to proceed when allocation or efficiency data for a country are unavailable.
The calculation pipeline in the [PFUDatabase][link-pfudatabase] packge 
allows allocation and efficiency data
for any country and year to be added at any time to improve the database.

@@

| Package | Purpose |
|:--------|:--------|
| [IEATools][link-ieatools]             | Converts [IEA WEEB][link-ieaweeb] data to RUVY matrices in [matsindf][link-matsindf] data frames |
| [MWTools][link-mwtools]               | Converts [ILO][link-ilo] and [FAO][link-fao] data to RUVY matrices of human and animal muscle work in [matsindf][link-matsindf] data frames |
| [Recca][link-recca]                   | Performs `R` energy conversion chain analysis |
| [PFUSetup][link-pfusetup]             | Identifies input and output locations for the [PFUDatabase][link-pfudatabase] and [PFUAggDatabase][link-pfuaggdatabase] pipelines |
| [PFUDatabase][link-pfudatabase]       | Provides a [targets][link-targets] pipeline to create a data frame of RUVY matrices |
| [PFUAggDatabase][link-pfuaggdatabase] | Provides a [targets][link-targets] pipeline to aggregate RUVY matrices |

Input data for the PFU database can be found in @Marshall:2023aa.
Access to the PFU database can be obtained via correspondence 
with author [PEB](mailto:P.E.Brockway@leeds.ac.uk)^[Because the PFU Database contains
primary and final energy [IEA WEEB][link-ieaweeb] data, 
access to the PFU Database
is restricted to those who have access to [IEA WEEB][link-ieaweeb] data.].
